DaLi: Database as a Library

### 2017-02-20 paper/motivation.tex
Counter eg modified.

\caption{\C{Counter} library in OCaml}
\label{fig:counter-adt}
\end{subfigure}
\begin{subfigure}[b]{0.56\textwidth}
  \begin{ocaml}
    module Counter: sig
      type t
      type eff = t -> t
      val add: int -> t-> eff
      val mult: int -> t -> eff
      val read: t -> int
    end = struct
      type t = int
      type eff = t -> t
      let add x v = fun v' -> v' + (abs x)
      let mult x v = 
        fun v' -> v' + v * (abs x - 1) 
      let read v = v
    end
  \end{ocaml}

\caption{\C{Counter} library re-engineered for op-based replication}
\label{fig:counter-rdt}
\end{subfigure}
% \begin{minipage}{0.62\textwidth}
%   \begin{ocaml}
%     module Counter = struct
%       include Counter
%       type t = counter [@@deriving persistence]
%       let merge old v1 v2 = old + (v1 - old) 
%                                 + (v2 - old)
%     end
%   \end{ocaml}
% \end{minipage}

@@

The counter implementation of Fig.~\ref{fig:counter-adt} serves the
application well as long as it is used in-memory on a single machine.
However, as the application grows in complexity and scales beyond a
single machine, its state may also need to be replicated and
persisted. Unfortunately, replication doesn't come for free, and
existing techniques require significant re-engineering of the
application and its libraries to support replication. A method often
adopted~\cite{crdts,gotsman-popl16} is to re-define the operations to
return \emph{effects} instead of values.  For instance, a \C{Counter}
operation returns a function of type $\C{Counter.t} \rightarrow
\C{Counter.t}$, which is then applied to the counter values at all
replicas to realize its effect.  Fig.~\ref{fig:counter-rdt} shows the
\C{Counter} library with operations re-defined to returns effects.
Such re-engineering has to be done carefully, lest an operation may
may end up causing an unintended effect at a remote replica. For
instance, while the effect of a \C{mult x v} operation may intuitively
seem to be \C{fun v' -> (abs x) * v'}, it is incorrect for it affects
different replicas differently, depending on the local counter value.
Fig.~\ref{fig:counter-rdt} therefore defines the result of \C{mult x
v} as an effect that increments the counter value by \C{v * (abs x -
1)} uniformly at all replicas. Likewise, accidentally defining the
effect of an \C{add x v} operation as \C{fun v' -> (abs x) + v}
results in replacing the counter value at a remote replica with the
one at origin replica, thereby losing updates at the remote replica.
Thus, reasoning about the correctness of effects often involves
appeals to system-specific artifacts, such as replicas, which don't
normally manifest in the program.  While checking the effects for
commutativity may help catch some of these oversights, commutativity
is a non-local property, and automatic verification is difficult in
general. Besides the possibility of new bugs, re-engineering libraries
to return effects has obvious software engineering problems in the
sense that it breaks backwards compatibility with the existing code.

>>>>

\caption{\C{Counter} library in OCaml}
\label{fig:counter-adt}
\end{subfigure}
\begin{subfigure}[b]{0.56\textwidth}
  \begin{ocaml}
    module Counter: sig
      type t
      type eff 
      val add: int -> t-> eff
      val mult: int -> t -> eff
      val apply: eff -> t -> t
      val read: t -> int
    end = struct
      type t = int
      type eff = Add of int
      let add x v = Add (abs x)
      let mult x v = Add (v * (abs x - 1))
      let apply (Add x) v = x + v
      let read v = v
    end
  \end{ocaml}

\caption{\C{Counter} library re-engineered for effect-based replication}
\label{fig:counter-rdt}
\end{subfigure}
% \begin{minipage}{0.62\textwidth}
%   \begin{ocaml}
%     module Counter = struct
%       include Counter
%       type t = counter [@@deriving persistence]
%       let merge old v1 v2 = old + (v1 - old) 
%                                 + (v2 - old)
%     end
%   \end{ocaml}
% \end{minipage}

@@

The counter implementation of Fig.~\ref{fig:counter-adt} serves the
application well as long as it is used in-memory on a single machine.
However, as the application grows in complexity and scales beyond a
single machine, its state may also need to be replicated and
persisted. Unfortunately, replication doesn't come for free, and
existing techniques require significant re-engineering of the
application and its libraries to support replication. A method often
adopted~\cite{crdts, pldi15, gotsman-popl16} is to re-define the operations to
return \emph{effects} instead of values.  An \emph{effect} is a tag that
identifies the operation to be applied uniformly at all
replicas to incorporate the effects of the original operaton. For
instance, \C{Counter.add x v} operation returns \C{Add x} effect,
which, when applied at a replica (see \C{apply}), adds \C{x} to the 
local counter value.  Fig.~\ref{fig:counter-rdt} shows the
\C{Counter} library with operations re-defined to returns effects.
Such re-engineering has to be done carefully, lest an operation may
may end up causing an unintended effect at a remote replica. For
instance, one may incorrectly define \C{mult x v} operation as
generating a \C{Mult x} effect, for it is easy to overlook the fact
that \C{apply (Mult x)} affects different replicas differently
depending on the counter value. Thus, reasoning about the correctness
of effects often involves appeals to system-specific artifacts, such
as replicas, which don't normally manifest in the program.  While
checking the effects (or rather their application via \C{apply}) for
commutativity may help catch some of these oversights, commutativity
is a non-local property, and automatically verifying it is difficult.
Besides introducing the possibility of new bugs, re-engineering
libraries to return effects has obvious software engineering problems
in the sense that it breaks backwards compatibility with the existing
code.

Another downside with effect-generating replicated data type libraries
is that they don't compose. For instance, let us say that an
application wants to use two replicated counters, $c_1$ and $c_2$,
bound by an invariant $c_2 \ge c_1$, which the application duly the
invariant when updating $c_2$ and $c_1$. Nonetheless, it may still
witness anamolous states that violate the invariant because updates to
$c_1$ and $c_2$ are applied independently in any order. For example,
if a replica increments both the counters by 1 each, the \C{read} at
another replica, where $(c_1,c_2) = (2,2)$ may witness the increment
to $c_1$, but not $c_2$, thus violating the invariant. The intention
to atomically apply both the effects cannot be captured in the absence
of external mechanisms to compose effects (such as effect-based
transactions).
